How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-29 docs/PhysicsPaper/StokesErosion.tex
NM

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium \cite{souzy2020velocity}. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a particular eroded geometry with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(-1,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate the configuration of eroded bodies by 90 degrees to compute a transverse tortuosity $T_2$.

@@

With the numerical methods in place, we now present results on how fluid-mechanical erosion alters porous-media properties over time. We first discuss the results from a single simulation and then generalize to statistical results from ensembles of simulations.

@@

As given in~\eqref{eq:anis}, the ratio of longitudinal to transverse permeability defines the anisotropy of the medium, $\anis = k_{11} / k_{22}$, which is plotted in Figure~\ref{fig3}(b) over the course of the simulation (solid curve). Early in the simulation the anisotropy is nearly one, as the initial configuration of randomly placed circles has no preferred flow direction. The anisotropy then increases as heterogeneous rates of erosion promote longitudinal flow over transverse flow. The anisotropy peaks at a value of $\anis \approx 4$ later in the simulation. At very late times, the anisotropy approaches unity again as the bodies completely vanish and return the system to a state of no preferred flow direction.

>>>>

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium \cite{souzy2020velocity}. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a partially-eroded configuration of bodies with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(-1,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate the configuration of eroded bodies by 90 degrees to compute a transverse tortuosity $T_2$.

@@

With the numerical methods in place, we now present results on how fluid-mechanical erosion alters porous-media properties over time. We first discuss the results from a single simulation and then generalize to statistical analysis of ensembles of simulations.

@@

As given in~\eqref{eq:anis}, the ratio of longitudinal to transverse permeability defines the anisotropy of the medium, $\anis = k_{11} / k_{22}$, which is plotted in Figure~\ref{fig3}(b) over the course of the simulation (solid curve). Early on, the anisotropy is nearly one, as the initial configuration of randomly placed circles has no preferred flow direction. The anisotropy then increases as heterogeneous rates of erosion promote longitudinal flow over transverse flow. The anisotropy peaks at a value of $\anis \approx 4$ later in the simulation. At very late times, the anisotropy approaches unity again as the bodies completely vanish and return the system to a state of no preferred flow direction.
